---
title: "Week 10: t distribution and t-tests"
author: 'Stat 201: Statistics I'
date: "April 14, 2019"
output: 
    beamer_presentation:
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      fig.width = 4, fig.height = 1.5, fig.align = "center")

library(ggplot2)

set.seed(42)
```

```{r}
normal.plot <- function(x = NULL,
                        x.ticks.sd = (-1:1)*2,
                        mu = 0, sigma = 1,
                        less.x = NULL,
                        less.x.2 = NULL,
                        greater.x = NULL,
                        greater.x.2 = NULL,
                        between.x = NULL,
                        fill.color = "cadetblue",
                        less.color = NULL,
                        less.color.2 = NULL,
                        greater.color = NULL,
                        greater.color.2 = NULL,
                        between.color = NULL) {
    if (is.null(x)){
        # If no x values given, default to -3.1 to 3.1 SDs
        l.bound <- mu - 3 * sigma
        u.bound <- mu + 3 * sigma
        
        x <- seq(l.bound, u.bound, length=500)
    } 

    x.ticks <- mu + x.ticks.sd * sigma
    x.ticks <- x.ticks[x.ticks>=min(x) & x.ticks<=max(x)]        
    
    # The density function for specified distribution
    dense.func <- function(x){
        dnorm(x, mean=mu, sd=sigma)
    }
    
    # y values are the density curve
    y <- dense.func(x)
    
    # Data frame for the plot
    plot.df <- data.frame(x=x, y=y)
    
    # Define the plot
    g <- ggplot(plot.df, aes(x=x, y=y))
    
    # Display range less than value?
    if (!is.null(less.x)){
        if (is.null(less.color)){
            less.color <- fill.color
        }
        
        area.less.x <- c(x[x<= less.x], less.x)
        plot.less.df <- data.frame(x=area.less.x, y=dense.func(area.less.x))
        
        g <- g + geom_area(data=plot.less.df, fill=less.color, alpha=.6)
        g <- g + geom_segment(aes(x=less.x, y=0, xend=less.x, yend=dense.func(less.x)),
                              color=less.color)
        
        x.ticks <- c(x.ticks, less.x)
    }
    
    if (!is.null(less.x.2)){
        if (is.null(less.color.2)){
            less.color.2 <- fill.color
        }
        
        area.less.x.2 <- c(x[x<= less.x.2], less.x.2)
        plot.less.df.2 <- data.frame(x=area.less.x.2, y=dense.func(area.less.x.2))
        
        g <- g + geom_area(data=plot.less.df.2, fill=less.color.2, alpha=.6)
        g <- g + geom_segment(aes(x=less.x.2, y=0, xend=less.x.2, yend=dense.func(less.x.2)),
                              color=less.color.2)
        
        x.ticks <- c(x.ticks, less.x.2)
    }

    # Display range greater than value?
    if (!is.null(greater.x)){
        if (is.null(greater.color)){
            greater.color <- fill.color
        }

        area.greater.x <- c(greater.x, x[x >= greater.x])
        plot.greater.df <- data.frame(x=area.greater.x, y=dense.func(area.greater.x))

        g <- g + geom_area(data=plot.greater.df, fill=greater.color, alpha=.6)
        g <- g + geom_segment(aes(x=greater.x, y=0, 
                                  xend=greater.x, yend=dense.func(greater.x)),
                              color=greater.color)
        
        x.ticks <- c(x.ticks, greater.x)
    }

    if (!is.null(greater.x.2)){
        if (is.null(greater.color.2)){
            greater.color.2 <- fill.color
        }

        area.greater.x.2 <- c(greater.x.2, x[x >= greater.x.2])
        plot.greater.df.2 <- data.frame(x=area.greater.x.2, y=dense.func(area.greater.x.2))

        g <- g + geom_area(data=plot.greater.df.2, fill=greater.color.2, alpha=.6)
        g <- g + geom_segment(aes(x=greater.x.2, y=0, 
                                  xend=greater.x.2, yend=dense.func(greater.x.2)),
                              color=greater.color.2)
        
        x.ticks <- c(x.ticks, greater.x.2)
    }

    # Display range between values?
    if (!is.null(between.x)){
        if (is.null(between.color)){
            between.color <- fill.color
        }

        area.between.x <- c(between.x[1], 
                            x[x >= between.x[1] & x <= between.x[2]],
                            between.x[2])
        plot.between.df <- data.frame(x=area.between.x, y=dense.func(area.between.x))

        g <- g + geom_area(data=plot.between.df, fill=between.color, alpha=.6)
        g <- g + geom_segment(aes(x=between.x[1], y=0, 
                                  xend=between.x[1], yend=dense.func(between.x[1])),
                              color=between.color)
        g <- g + geom_segment(aes(x=between.x[2], y=0, 
                                  xend=between.x[2], yend=dense.func(between.x[2])),
                              color=between.color)
        
        x.ticks <- c(x.ticks, between.x)
    }

    x.ticks <- sort(x.ticks)
    x.ticks.labels <- paste(round(x.ticks,2))
                            
    # Draw density curve
    g <- g + geom_line(size=1)
    g <- g + theme_bw() + xlab("") + ylab("Density")
    g <- g + scale_x_continuous(breaks=sort(x.ticks), minor_breaks = NULL,
                                labels=x.ticks.labels)
    
    g
}

binomial.plot <- function(x = NULL,
                        x.ticks = NULL,
                        n = 10, p = 0.5,
                        less.x = NULL,
                        less.x.2 = NULL,
                        greater.x = NULL,
                        greater.x.2 = NULL,
                        between.x = NULL,
                        fill.color = "cadetblue",
                        less.color = NULL,
                        less.color.2 = NULL,
                        greater.color = NULL,
                        greater.color.2 = NULL,
                        between.color = NULL) {
    if (is.null(x)){
        # If no x values given, default to 0 to n
        l.bound <- - 0.51
        u.bound <- n + 0.5 
        
        x <- seq(l.bound, u.bound, length=500)
    } 
    
    if (is.null(x.ticks)){
        x.ticks <- seq(round(min(x+0.01)), round(max(x)), by=2)
    }

    # The density function for specified distribution
    dense.func <- function(x){
        dbinom(round(x), size=n, prob=p)
    }
    
    # y values are the density curve
    y <- dense.func(x)
    
    # Data frame for the plot
    plot.df <- data.frame(x=x, y=y)
    
    # Define the plot
    g <- ggplot(plot.df, aes(x=x, y=y))
    
    # Display range less than value?
    if (!is.null(less.x)){
        if (is.null(less.color)){
            less.color <- fill.color
        }

        area.less.x <- x[x<= less.x + 0.5]
        plot.less.df <- data.frame(x=area.less.x, y=dense.func(area.less.x))
        
        g <- g + geom_area(data=plot.less.df, fill=less.color, alpha=.6)
        g <- g + geom_segment(aes(x=less.x+0.5, y=0, 
                                  xend=less.x+0.5, yend=dense.func(less.x+0.5)),
                              color=less.color)
        
        x.ticks <- c(x.ticks, less.x)
    }
    

    if (!is.null(less.x.2)){
        if (is.null(less.color.2)){
            less.color.2 <- fill.color
        }

        area.less.x.2 <- x[x<= less.x.2 + 0.5]
        plot.less.df.2 <- data.frame(x=area.less.x.2, y=dense.func(area.less.x.2))
        
        g <- g + geom_area(data=plot.less.df.2, fill=less.color.2, alpha=.6)
        g <- g + geom_segment(aes(x=less.x.2+0.5, y=0, 
                                  xend=less.x.2+0.5, yend=dense.func(less.x.2+0.5)),
                              color=less.color.2)
        
        x.ticks <- c(x.ticks, less.x.2)
    }

    # Display range greater than value?
    if (!is.null(greater.x)){
        if (is.null(greater.color)){
            greater.color <- fill.color
        }

        area.greater.x <- x[x >= greater.x + 0.5]
        plot.greater.df <- data.frame(x=area.greater.x, y=dense.func(area.greater.x))

        g <- g + geom_area(data=plot.greater.df, fill=greater.color, alpha=.6)
        g <- g + geom_segment(aes(x=greater.x + 0.5, y=0, 
                                  xend=greater.x + 0.5, yend=dense.func(greater.x+0.5)),
                              color=greater.color)
        
        x.ticks <- c(x.ticks, greater.x)
    }

    if (!is.null(greater.x.2)){
        if (is.null(greater.color.2)){
            greater.color.2 <- fill.color
        }

        area.greater.x.2 <- x[x >= greater.x.2 + 0.5]
        plot.greater.df.2 <- data.frame(x=area.greater.x.2, y=dense.func(area.greater.x.2))

        g <- g + geom_area(data=plot.greater.df.2, fill=greater.color.2, alpha=.6)
        g <- g + geom_segment(aes(x=greater.x.2 + 0.5, y=0, 
                                  xend=greater.x.2 + 0.5, yend=dense.func(greater.x.2+0.5)),
                              color=greater.2.color)
        
        x.ticks <- c(x.ticks, greater.x.2)
    }

    # Display range between values?
    if (!is.null(between.x)){
        if (is.null(between.color)){
            between.color <- fill.color
        }

        area.between.x <- c(between.x[1] + 0.5, 
                            x[x > between.x[1] + 0.5 & x <= between.x[2] + 0.5])
        plot.between.df <- data.frame(x=area.between.x, y=dense.func(area.between.x))

        g <- g + geom_area(data=plot.between.df, fill=between.color, alpha=.6)
        g <- g + geom_segment(aes(x=between.x[1] + 0.5, y=0, 
                                  xend=between.x[1] + 0.5, yend=dense.func(between.x[1] + 0.5)),
                              color=between.color)
        g <- g + geom_segment(aes(x=between.x[2] + 0.5, y=0, 
                                  xend=between.x[2] + 0.5, yend=dense.func(between.x[2]+0.5)),
                              color=between.color)
        
        x.ticks <- c(x.ticks, between.x)
    }

    x.ticks <- sort(x.ticks)
    x.ticks.labels <- paste(round(x.ticks,4))
                            
    # Draw density curve
    g <- g + geom_step(size=1, direction = "vh")
    g <- g + theme_bw() + xlab("") + ylab("Density")
    g <- g + scale_x_continuous(breaks=sort(x.ticks), minor_breaks = NULL,
                                labels=x.ticks.labels)
    
    g
}

```

```{r}
hgts <- read.csv("../../Data/metro_hgts_pop.csv")
s <- sample(hgts$x, 40)
```

## Tests for proportions

Proportion tests are conducted with the `prop.test` function.

`prop.test(x, n, p, alternative, correct=FALSE)`

- `x` is the number of successes. Can be a single number for one sample tests, or a vector for two sample tests.
- `n` is the number of trials. Must be the same length as the `x` parameter.
- `p` is the null proportion to test against.
- `alternative` is the form of the alternative hypothesis ("two.sided", "less", "greater").
- `correct` is whether to apply a continuity correction. Defaults to `TRUE`, but should be set to `FALSE` to match results from the formulas.


# One sample proportion test

- A sample has 62 successes out of 212 trials. Test whether the population proportion is less than 40\%.

    ```{r, echo=T, eval=F}
    prop.test(62, 212, 0.4, alternative="less", correct=FALSE)
    ```
    \vspace{-.15in}
    ```{r, echo=F, eval=T}
    prop.test(62, 212, 0.4, alternative="less", correct=FALSE)
    ```

# Two sample proportion test

- A sample has 62 successes out of 212 trials and another has 59 successes out of 173 trials. Test whether the population proportions are the same.

    ```{r, echo=T, eval=F}
    prop.test(c(62, 59), c(212, 173), correct=FALSE)
    ```
    \vspace{-.15in}
    ```{r, echo=F, eval=T}
    prop.test(c(62, 59), c(212, 173), correct=FALSE)
    ```

# Goodness-of-fit tests

Goodness-of-fit tests are conducted with the `chisq.test` function.

`chisq.test(x, p)`

- `x` is a vector of category counts.
- `p` is a vector of probabilities (same length as `x`). If not included, defaults to equal probabilities.

# Goodness-of-fit test example

- A 6-sided die is rolled 100 times. The results are 24, 16, 13, 16, 17, 14 for each number respectively. Perform a goodness-of-fit test of equal probabilities.

    ```{r, echo=T, eval=F}
    chisq.test(c(24, 16, 13, 16, 17, 14))
    ```
    \vspace{-.15in}
    ```{r, echo=F, eval=T}
    chisq.test(c(24, 16, 13, 16, 17, 14))
    ```

# Goodness-of-fit test example

- Test whether a sample with category counts of 13, 48 and 31 matches an expected distribution of 25%, 50% and 25%.

    ```{r, echo=T, eval=F}
    chisq.test(c(13, 48, 31), p=c(.25, .5, .25))
    ```
    \vspace{-.15in}
    ```{r, echo=F, eval=T}
    chisq.test(c(13, 48, 31), p=c(.25, .5, .25))
    ```

# Tests for independence


Tests for independence are alse conducted with the `chisq.test` function.

`chisq.test(x, correct)`

- `x` is a matrix representing a contingency table.
- `correct` is whether to apply continuity correction. Should be set to `FALSE`.

# Matrices

A matrix is created with the `matrix` function.

`matrix(data, n.row)`

- `data` is a vector containing the values to populate the matrix. The order of the values should be 1st column - top to bottom, 2nd column - top to bottom, etc.

- `n.row` is the number of rows of the matrix. The number of columns will then be deterined by the length of `data`


# Matrix example

- Save the contingency table as a matrix.
    
    \begin{tabular}{c | c c}
    & 1A & 1B\\
    \hline
    2A & 10 & 30\\
    2B & 20 & 40\\
    \end{tabular}

    ```{r, echo=T, eval=F}
    x <- matrix(c(10, 20, 30, 40), nrow=2)
    x
    ```
    \vspace{-.15in}
    ```{r, echo=F, eval=T}
    x <- matrix(c(10, 20, 30, 40), nrow=2)
    x
    ```
    
# Test of independence example

- Perform a test of independence for the sample in the table below.

    \begin{tabular}{c | c c}
    & Positive & Negative\\
    \hline
    Cancer & 74 & 13\\
    No cancer & 26 & 887\\
    \end{tabular}

    ```{r, echo=T, eval=F}
    x <- matrix(c(74, 26, 13, 887), nrow=2)
    chisq.test(x, correct=FALSE)
    ```
    \vspace{-.15in}
    ```{r, echo=F, eval=T}
    x <- matrix(c(74, 26, 13, 887), nrow=2)
    chisq.test(x, correct=FALSE)
    ```
