\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}

\begin{flushleft}
\section*{Group Work - Week 2}
\paragraph{1} Consider rolling a fair six-sided die.
\begin{enumalpha}
\item Let event $A$ be rolling an even number. What is a trial for this scenario? What is the sample space? Is $A$ a simple event? What is $P(A)$? What is $\bar A$, the complement of $A$? What is $P(\bar A)$? Is $A$ unlikely? Is $A$ unusual?
\vspace{2.5in}
\item Let event $A$ be rolling an even number. Let event $B$ be rolling a 3. Are events $A$ and $B$ disjoint? What is $P(A \text{ or } B)$?
\vspace{2.5in}
\item Consider rolling a die twice. Let event $A$ be getting an even number on the first roll. Let event $B$ be getting 5 or more on the second roll. Are events $A$ and $B$ independent? What is $P(A \text { and } B)$?
\end{enumalpha}

\newpage
\paragraph{2} Consider a standard deck of playing cards... 52 cards, 4 suits of 13 cards each, 3 cards of each suit are face cards, 2 suits are black (clubs and spades) and 2 are red (hearts and diamonds).
\begin{enumalpha}
\item Let event $A$ be drawing a random card that is a diamond. What is a trial for this scenario? What is the sample space? Is $A$ a simple event? What is $P(A)$? What is $\bar A$, the complement of $A$? What is $P(\bar A)$? Is $A$ unlikely? Is $A$ unusual?
\vspace{2.5in}
\item Let event $A$ be drawing a random card that is a diamond. Let event $B$ be drawing a random card that is a face card. Are events $A$ and $B$ disjoint? What is $P(A \text{ or } B)$?
\vspace{2.5in}
\item Consider drawing three cards. Let event $A$ be the first card is a heart. Let event $B$ be the second card is a club. Let event $C$ be the third card is black. Are events $A$, $B$ and $C$ independent? What is $P(A \text { and } B \text { and } C)$?

\end{enumalpha}

\newpage
\paragraph{3} The data set ``hair\_eye.csv" on D2L contains the hair and eye colors, as well as sex, of a sample of statistics students. Below is a table showing the distributions of students by eye color and gender.\\
\medskip
\renewcommand{\arraystretch}{1}
{\centering
\begin{tabular}{ r| c c c c}
& \multicolumn{4}{c}{Eye color}\\
Gender & Blue & Brown & Green & Hazel\\
\hline
 Female &  114   &  122  &  31  &  46\\
 Male   & 101  &  98  &  33  &  47\\
\end{tabular}
\par}
\begin{enumalpha}
\item Let event $A$ be a randomly selected student having green eyes. What is a trial for this scenario? What is the sample space? Is $A$ a simple event? What is $P(A)$? What is $\bar A$, the complement of $A$? What is $P(\bar A)$? Is $A$ unlikely? Is $A$ unusual?
\vspace{2.25in}
\item Let event $A$ be a randomly selecting a student with brown or blue eyes. Let event $B$ be a randomly selecting a female student. Are events $A$ and $B$ disjoint? What is $P(A \text{ or } B)$?
\vspace{2.25in}
\item Consider randomly selecting two students. Let event $A$ be the first student has blue eyes. Let event $B$ be the second student has hazel eyes. Are events $A$ and $B$ independent? What is $P(A \text { and } B)$?

\end{enumalpha}



\end{flushleft}
\end{document}