\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}

\begin{flushleft}
\section*{Group Work - Week 5}

\paragraph{1} The file ``\verb+heights.csv+" on D2L has heights in inches of a sample of 25 men and 25 women.\\ \medskip


\begin{enumalpha}
\item Create 5 number summaries and boxplots for both men and women.
\vspace{1.75in}

\item Consider a man who is 68.5 inches tall and a woman who is 63.25 inches tall. What are the z-scores for height for both of these individuals? Who is taller relative to their gender?
\vspace{1.75in}

\item What percentile is 65 inches for a man? What height is the 80th percentile for women?

\end{enumalpha}

\newpage
\paragraph{2} Suppose you encounter the following game at a school carnival: A wheel with colored spokes is spun and the wheel comes to rest with a pointer pointing at one of the spokes. If you land on a gold spoke, you get 10 tickets. You get 4 tickets for landing on blue and 2 tickets for landing on red. If you land on black, you get nothing. There are 9 black spokes, 6 red spokes, 4 blue and 1 gold spoke. The game cost three tickets to play.

\begin{enumalpha}
\item Create a probability distribution which describes this game.
\vspace{1.75in}

\item How many tickets should you expect to win? That is, what is the expected value of the game? What is the standard deviation of tickets won?
\vspace{1.75in}

\item Is this a fair game? If your goal was to increase you tickets, what is the maximum number of tickets you should pay for the game?
\end{enumalpha} 
\end{flushleft}
\end{document}