\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}

\begin{flushleft}
\section*{Group Work - Week 6}
\paragraph{1} The best player on the Metro State basketball team successfully makes 85\% of her free throws. During a typical game she attempts 15 free throws. (If she attempts a different number of free throws, it is not a typical game.)
\begin{enumalpha}
\item Is the number of free throws she makes in a typical game a proper binomial random variable? What are the values of $n$, $p$ and $q$?

\vspace{1.8in}
\item What is the expected number of free throws made per game? What is the standard deviation?

\vspace{1.8in}
\item What is the probability she makes all her free throws in a typical game? Is that unusual? What is the probability she makes at least 12 free throws? Less than 12?

\vspace{1.8in}
\item What are unusual numbers of free throws made per game?
\end{enumalpha}

\newpage

\paragraph{2} Eleanor is taking two classes, statistics and ethics. After the midterm exams, Eleanor got 86 on her statistics midterm and she scored better than 88\% of the class on her ethics midterm. The statistics midterm scores were approximately normally distributed with a mean of 84 and a standard deviation of 5.2. The ethics midterm scores were approximately normally distributed with a mean of 71.5 and a standard deviation of 9.8. 

\begin{enumalpha}
\item What is the probability a randomly selected student did worse than Eleanor on the statistics midterm? That is, what is $P(X < 86)$? Or, what percentile is Eleanor's statistics grade?

\vspace{2in}
\item What score did Eleanor get on her ethics midterm? That is, for what $x$ is $P(X < x) = 0.88$? 

\vspace{2in}
\item Calculate the z-scores for Eleanor's two midterm grades. In which class did she do better?
\end{enumalpha}

\end{flushleft}
\end{document}