\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}

\begin{flushleft}
\section*{Group Work - Week 9}

\paragraph{1} The data file ``bears.csv" on D2L contains measurements of a random sample of bears from a national park. Harsh winters can be hard on a bear population, especially older bears. Park officials want to know if the mean bear age is different than the usual mean of 55 months.

\begin{enumalpha}
\item What are the null and alternative hypotheses for a test on this claim? Is this a one-sided or two-sided test? Is the claim represented by the null or alternative hypothesis?
\vspace{2.5in}
\item Using the data set, conduct a test at the $\alpha = 0.05$ level of significance of the claim that the bear population has a different mean age of 55 months. Be sure to state your conclusion in the context of the question.
\end{enumalpha}

\newpage
\paragraph{2} A manufacturer of flash drives wants to know if there is a difference in the reliability of their drives used in extreme conditions. A sample of 15 drives used in cold conditions ($< 32 $\textdegree F) had a mean lifespan of 41.9 months with a standard deviation of 6.3. A sample of 15 drives used in hot conditions ($> 99 $\textdegree F) had a mean lifespan of 38.4 months with a standard deviation of 5.9. Assume lifespans of flash drive are normally distributed.
\begin{enumalpha}
\item What are the null and alternative hypotheses for a test on this claim? Is this a one-sided or two-sided test? Are these independent or dependent samples? Are the requirements for a hypothesis test satisfied?
\vspace{2.25in}
\item Conduct an hypothesis test at the $\alpha = 0.05$ level of significance. Be sure to state your conclusion in the context of the question.
\end{enumalpha}

\newpage
\paragraph{3} Researchers are interested in whether meditation can lower blood pressure in people that have high blood pressure. They conduct a study on 45 patients with high blood pressure (systolic blood pressure $> 20$), measuring their systolic blood pressure at baseline and after 30 minutes of meditation. The file ``meditation\_bp.csv" on D2L contains the data.
\begin{enumalpha}
\item What are the null and alternative hypotheses for a test on this claim? Is this a one-sided or two-sided test? Are these independent or dependent samples? Are the requirements for a hypothesis test satisfied?
\vspace{2.25in}
\item Conduct an hypothesis test at the $\alpha = 0.01$ level of significance. Be sure to state your conclusion in the context of the question.
\vspace{2.25in}
\end{enumalpha}



\end{flushleft}
\end{document}