\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}
\flushleft

\section*{Homework - Week 8}
\begin{itemize}
\item \bt{From OpenIntro Statistics, chapter 4: } 4.17 (a-b), 4.18 (a-b), 4.19, 4.20, 4.21 (a-c)

\item For the following scenarios, give the required sample size.
\begin{enumalpha}
\item A university wants to know how much time their students are spending on homework each week to within plus or minus 15 minutes (0.25 hours) with 95\% confidence. Previous studies have suggested the standard deviation for weekly study times is about 3 hours.

\item The university would also like to know what proportion of their students utilize university resources, like tutoring help, to within 3.5\% with 99\% confidence. Previous years found about 35\% of students used available resources, but the university thinks that has increased recently.

\item The student association at the university would like to know what proportion of professors offer extra credit in their classes. They plan to gather a random sample of syllabuses and see which mention extra credit. They have no reasonable estimate as to what that proportion might be. They want an estimate within 5\% at a 90\% confidence level.

\item The university bookstore would like to know how much the average student spends on textbooks each semester to within \$10 (plus or minus) with 95\% confidence. From experience, they think that a standard deviation of \$30 is a reasonable estimate.
\end{enumalpha}

\end{itemize}

All submitted files should be PDFs.

\end{document}