\documentclass{article}

\usepackage{shyne}

% document format
\topmargin 0in
\oddsidemargin 0in
\evensidemargin 0in
\headheight 0in
\headsep 0in
\topskip 0in
\textheight 9in
\textwidth 6.5in
\linespread{1.3}

\begin{document}
\flushleft

\section*{Homework - Week 8}
\begin{itemize}
\item \bt{4.17 (a-b)}  Write the null 
and alternative hypotheses in words and then symbols for each of the following 
situations.
\begin{enumalpha}
\item New York is known as ``the city that never sleeps". A random sample of 25 
New Yorkers were asked how much sleep they get per night. Do these data provide 
convincing evidence that New Yorkers on average sleep less than 8 hours a night?\\
\medskip
\bt{Null hypothesis: New Yorkers sleep 8 hours a night on average}\\
\bt{Alternative hypothesis: New Yorkers sleep less than 8 hours a night on average}\\
$\bv{H_0: \mu = 8}$\\
$\bv{H_a: \mu < 8}$
\vspace{0.5in}

\item Employers at a firm are worried about the effect of March Madness, a 
basketball championship held each spring in the US, on employee productivity. 
They estimate that on a regular business day employees spend on average 15 
minutes of company time checking personal email, making personal phone calls, 
etc. They also collect data on how much company time employees spend on such non-
business activities during March Madness. They want to determine if these data 
provide convincing evidence that employee productivity decreases during March 
Madness.
\\
\medskip
\bt{Null hypothesis: Employees spend 15 minutes per day on personal matters, on average}\\
\bt{Alternative hypothesis: Employees spend more than 15 minutes per day on personal matters, on average}\\
$\bv{H_0: \mu = 15}$\\
$\bv{H_a: \mu >15}$
\vspace{0.5in}
\end{enumalpha}
 

\item \bt{4.18 (a-b)}  Write the null and alternative hypotheses in words and using symbols 
for each of the following situations.
\begin{enumalpha}
\item Since 2008, chain restaurants in California have been required to display 
calorie counts of each menu item. Prior to menus displaying calorie counts, the 
average calorie intake of diners at a restaurant was 1100 calories. After 
calorie counts started to be displayed on menus, a nutritionist collected data 
on the number of calories consumed at this restaurant from a random sample of 
diners. Do these data provide convincing evidence of a difference in the average 
calorie intake of a diners at this restaurant?\\
\medskip
\bt{Null hypothesis: Diners consume 1100 calories on average}\\
\bt{Alternative hypothesis: Diners consume an amount different than 1100 calories on average}\\
$\bv{H_0: \mu = 1100}$\\
$\bv{H_a: \mu \ne 1100}$
\vspace{0.5in}

\item Based on the performance of those who took the GRE exam between July 1, 
2004 and June 30, 2007, the average Verbal Reasoning score was calculated to be 
462. In 2011 the average verbal score was slightly higher. Do these data provide 
convincing evidence that the average GRE Verbal Reasoning score has changed 
since 2004?\\
\medskip
\bt{Null hypothesis: The average GRE Verbal Reasoning score is 462}\\
\bt{Alternative hypothesis: The average GRE Verbal Reasoning score is not 462}\\
$\bv{H_0: \mu = 462}$\\
$\bv{H_a: \mu \ne 462}$
\vspace{0.5in}

\end{enumalpha}


\item \bt{4.19} A study suggests that the 
average college student spends 10 hours per week communicating with others 
online. You believe that this is an underestimate and decide to collect your 
own sample for a hypothesis test. You randomly sample 60 students from your 
dorm and find that on average they spent 13.5 hours a week communicating with 
others online. A friend of yours, who offers to help you with the hypothesis 
test, comes up with the following set of hypotheses. Indicate any errors you see.
\begin{align*}
H_0&: \bar{x} < 10~hours \\
H_A&: \bar{x} > 13.5~hours
\end{align*}\\
\medskip
\begin{itemize}
\item \bt{Hypotheses are statements about population parameters. The symbol $\bv{\mu}$ should be used instead of $\bv{\bar x}$. }

\item \bt{The null hypothesis should always be an equality statement, i.e $\bv{\bar x = 10}$}. 
\end{itemize}
\vspace{0.5in}




\item \bt{4.20} Exercise~4.16 presents the results 
of a 2006 - 2010 survey showing that the average age of women at first marriage 
is 23.44. Suppose a social scientist believes that this value has increased in 
2012, but she would also be interested if she found a decrease. Below is how she 
set up her hypotheses. Indicate any errors you see.
\begin{align*}
H_0&: \bar{x} = 23.44~years~old \\
H_A&: \bar{x} > 23.44~years~old
\end{align*}\\
\medskip
\begin{itemize}
\item \bt{Hypotheses are statements about population parameters. The symbol $\bv{\mu}$ should be used instead of $\bv{\bar x}$. }

\item \bt{Because the researcher is interested in increases or decreases, the alternative hypothesis should be $\bv{\bar x \ne 23.44}$}. 
\end{itemize}
\vspace{0.5in}



\item \bt{4.21 (a-c)} Exercise~4.13 
provides a 95\% confidence interval for the mean waiting time at an emergency 
room (ER) of (128 minutes, 147 minutes). Answer the following questions based on
this interval.
\begin{enumalpha}
\item A local newspaper claims that the average waiting time at this ER exceeds 
3 hours. Is this claim supported by the confidence interval? Explain your 
reasoning.\\
\medskip
\bt{No. 3 hours (180 minutes) is not in the reported confidence interval. Thus, there is no evidence that the true population mean is 3 hours.}\\
\vspace{0.5in}

\item The Dean of Medicine at this hospital claims the average wait time is 2.2 
hours. Is this claim supported by the confidence interval? Explain your 
reasoning.\\
\medskip
\bt{Maybe. 2.2 hours (132 minutes) is within the confidence interval, so there is not evidence that the claim is wrong. However, confidence intervals do not provide evidence for a particular value over another that is also in the interval.}\\
\vspace{0.5in}

\item Without actually calculating the interval, determine if the claim of the 
Dean from part (b) would be supported based on a 99\% confidence interval?\\
\medskip
\bt{Yes. A 99\% confidence interval will be wider than a 95\% confidence interval. Thus, a value that is within a 95\% CI will also be in a 99\% CI. }\\
\vspace{0.5in}

\end{enumalpha}


\item For the following scenarios, give the required sample size.
\begin{enumalpha}
\item A university wants to know how much time their students are spending on homework each week to within plus or minus 15 minutes (0.25 hours) with 95\% confidence. Previous studies have suggested the standard deviation for weekly study times is about 3 hours.\\
\medskip
$\ds \bv{n = \Paren{\frac{z_\alpha \times \sigma}{ME}}^2 = \Paren{\frac{1.96 \times 3}{0.25}}^2 = 553.1904 \to 554}$

\vspace{0.5in}

\item The university would also like to know what proportion of their students utilize university resources, like tutoring help, to within 3.5\% with 99\% confidence. Previous years found about 35\% of students used available resources, but the university thinks that has increased recently.\\
\medskip
$\ds \bv{n = \Paren{\frac{z_\alpha \times \sqrt{pq}}{ME}}^2 = \Paren{\frac{2.58 \times \sqrt{0.35 \times 0.65}}{0.035}}^2 = 1236.189 \to 1237 }$
\vspace{0.5in}

\item The student association at the university would like to know what proportion of professors offer extra credit in their classes. They plan to gather a random sample of syllabuses and see which mention extra credit. They have no reasonable estimate as to what that proportion might be. They want an estimate within 5\% at a 90\% confidence level.\\
\medskip
$\ds \bv{n = \Paren{\frac{z_\alpha \times \sqrt{pq}}{ME}}^2 = \Paren{\frac{1.645 \times \sqrt{0.5 \times 0.5}}{0.05}}^2 = 270.6025 \to 271 }$
\vspace{0.5in}

\item The university bookstore would like to know how much the average student spends on textbooks each semester to within \$10 (plus or minus) with 95\% confidence. From experience, they think that a standard deviation of \$30 is a reasonable estimate.\\
\medskip
$\ds \bv{n = \Paren{\frac{z_\alpha \times \sigma}{ME}}^2 = \Paren{\frac{1.96 \times 30}{10}}^2 = 34.5744 \to 35}$
\vspace{0.5in}
\end{enumalpha}

\end{itemize}

All submitted files should be PDFs.

\end{document}