---
title: "Homework - Week 4 (solution)"
author: "Michael Shyne"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1. Consider the builtin dataset `iris`.

    a. What is the structure of the `iris` data frame?
    
        ```{r}
        str(iris)
        ```
    
    b. Create a histogram of the `Sepal.Width` variable.
    
        ```{r}
        hist(iris$Sepal.Width)
        ```
    
    c. Create a histogram of the `Petal.Width` variable.

        ```{r}
        hist(iris$Petal.Width)
        ```

    d. For both histograms, does the data appear normally distributed? Are they skewed?
    
        **The sepal width histogram is approximately normal, with a slight right-skew.**
        
        **The petal width histogram is not normal. The data appear right-skewed, but since the distribution is bimodal defining a skew is not really appropriate.**
    
    e. For both histograms, does it appear that the data come from more than one populations?
    
        **The sepal width histogram does not show any evidence of coming from two populations.**
        
        **The petal width histogram is bimodal, suggesting the data come from two populations.**
    
    f. What is the mean and median of `Sepal.Width`? What is the variance and standard deviation?
    
        **Mean:**
        ```{r}
        mean(iris$Sepal.Width)
        ```
        
        **Variance and standard deviation:**
        ```{r}
        var(iris$Sepal.Width)
        sd(iris$Sepal.Width)
        ```
    
    g. What is the mean and median of `Petal.Width`? What is the variance and standard deviation?

        **Mean:**
        ```{r}
        mean(iris$Petal.Width)
        ```
        
        **Variance and standard deviation:**
        ```{r}
        var(iris$Petal.Width)
        sd(iris$Petal.Width)
        ```
    
    
2. Consider the builtin dataset `trees`. 

    a. What is the structure of the `trees` data frame?
        
        ```{r}
        str(trees)
        ```
        
    
    b. Create a histogram of the `Height` variable.
    
        ```{r}
        hist(trees$Height)
        ```
    
    c. Create a histogram of the `Volume` variable.
    
        ```{r}
        hist(trees$Volume)
        ```

    d. For both histograms, does the data appear normally distributed? Are they skewed?
    
        **The height histogram is approximately normal, with a slight left-skew.**
        
        **The volume histogram is not normal. It is heavily skewed to the right.** 

    e. For both histograms, does it appear that there are outliers in the data?
    
        **The height histogram does not display evidence of outliers.**
        
        **The volume histogram appears to have outliers.**
        
    f. What is the mean and median of `Height`? What is the variance and standard deviation?
    
        ```{r}
        height.mean <- mean(trees$Height)
        height.var <- var(trees$Height)
        ```
        
        **The mean tree height is `r height.mean`, the variance is `r height.var` and the standard deviation is the square root of variance or `r sqrt(height.var)`.**
    
    g. What is the mean and median of `Volume`? What is the variance and standard deviation?
    
        ```{r}
        # The round() function is helpful if you don't want
        #   to display many decimal places
        volume.mean <- round(mean(trees$Volume), 3)
        volume.var <- round(var(trees$Volume), 3)
        volume.sd <- round(sd(trees$Volume), 3)
        ```
        
        **The mean tree volume is (to 3 decimal places) `r volume.mean`, the variance is `r volume.var` and the standard deviation is `r volume.sd`.**

3. Load the dataset `bears.csv` from D2L.

    a. What is the structure of the `bears` data frame?
    
        ```{r}
        # Note: this assumes the `bears.csv` file is in the same folder as this
        #   markdown file (*.Rmd). If the file is located elsewhere, the read.csv() 
        #   must be passed a path to the file
        bears <- read.csv("bears.csv")
        
        str(bears)
        ```
    
    b. Create a frequency table for the variable `MONTH`. What is the mode, if any?
    
        ```{r}
        table(bears$MONTH)
        ```
    
        **Month 9 (September) has the highest frequency at 16, and thus is the mode.**
        
    c. Create a histogram of the `WEIGHT` variable.
    
        ```{r}
        hist(bears$WEIGHT)
        ```
    
    d. Is there distribution of `WEIGHT` data normal? Is it skewed? Are there outliers?
    
        **The weight histogram shows that the data are not normal. They are strongly right-skewed and there do appear to be outliers (weight > 500).**

    e. Based on your answers to part (d), do you expect the mean and median to be the same (or very close)? If not, which do you expect to be greater?
    
        **Because the data are skewed, the mean and median are likely not close to the same. Because the data are right-skewed with high value outliers, the mean will be pulled to a higher value. Thus, the mean should be greater than the median.**
    
    f. What is the mean and median of `WEIGHT`?
    
        **Mean:**
        ```{r}
        mean(bears$WEIGHT)
        ```
        
        **Median:**
        ```{r}
        median(bears$WEIGHT)
        ```
        
    g. Based on the histogram in part (c), what would you expect the mode to be, approximately?
    
        **The tallest bar in the histogram is for the range from 100 to 150. Thus, we would expect the mode to be in this range.**
        
    h. What is the mode, if any?

        ```{r}
        sort(table(bears$WEIGHT))
        ```
        
        **Technically, there are 6 modes, {140, 150, 166, 202, 204, 220}. In a practical sense, the fact that six values occur twice, while the rest occur once, is not very useful information. The peak of the histogram is a more useful estimate than the literal modes. This is often the case with continuous variables.**